Modelling state-switching behaviour in functional data with hidden Markov models

### 2020-08-31 Report/Sections/simulation.tex
Commit Before Andrew Access
Updates from Overleaf

To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT. The parameters used to generate the data are loosely based on those estimated in the case study (see able \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity. For example, we set $\Gamma$ such that the simulated data has approximately equal numbers of short and long dives. This allows meaningful conclusions to be made for both dive types. In addition, the number subdive states is set to $N^*=2$ and $\Ztwo$ is set to a scalar. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times. To assess the accuracy of uncertainty estimates, we also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model.

\subsection{Data Simulation}

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is gamma distributed, and the coarse-scale emission parameters $\Theta$ are
$$
    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
$$
%
After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
%
The observations $Z^*$ are generated using the following parameters:
%
\begin{gather*}
    \mu_A^{*(1)} = 0.0 s, \enspace \sigma_A^{*(1)} = 0.05s, \enspace \phi_A^{*(1)} = 0.99, \\
    %
    \mu_A^{*(2)} = 0.0 s, \enspace \sigma_A^{*(2)} = 0.1s, \enspace \phi_A^{*(2)} = 0.95, \\
    %
    \mu_W^{*(1)} = 10.10, \quad \sigma_W^{*(1)} = 3.18, \\
    %
    \mu_W^{*(2)} = 305.94, \quad \sigma_W^{*(2)} = 17.49.
\end{gather*}
%
Recall that $\Zone$ follows a CarHMM and is Normally distributed while $\Ztwo$ follows an HMM and is Gamma distributed. It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows this reconstructed sequence along with one realization of $Z^*$ for five dives of one simulated data set.

@@

\subsection{Simulation Results}

>>>>

To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT. The parameters used to generate the data are loosely based on those estimated in the case study (see able \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity. For example, we set $\Gamma$ such that the simulated data has approximately equal numbers of short and long dives. This allows meaningful conclusions to be made for both dive types. In addition, the number subdive states is set to $N^*=2$ and $\Ztwo_{t,t^*}$ is set to a scalar. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times. To assess the accuracy of uncertainty estimates, we also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model.

\subsection{Data simulation}

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is gamma distributed, and the coarse-scale emission parameters $\Theta$ are
$$
    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
$$
%
After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
%
The observations $Z^*$ are generated using the following parameters:
%
\begin{gather*}
    \mu_A^{*(1)} = 0.0 s, \enspace \sigma_A^{*(1)} = 0.05s, \enspace \phi_A^{*(1)} = 0.99, \\
    %
    \mu_A^{*(2)} = 0.0 s, \enspace \sigma_A^{*(2)} = 0.1s, \enspace \phi_A^{*(2)} = 0.95, \\
    %
    \mu_W^{*(1)} = 10.10, \quad \sigma_W^{*(1)} = 3.18, \\
    %
    \mu_W^{*(2)} = 305.94, \quad \sigma_W^{*(2)} = 17.49.
\end{gather*}
%
Recall that $\Zone_{t,t^*}$ follows a CarHMM and is Normally distributed while $\Ztwo_{t,t^*}$ follows an HMM and is Gamma distributed. It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows this reconstructed sequence along with one realization of $Z^*$ for five dives of one simulated data set.

@@

\subsection{Simulation results}
